---
title: "Midterm Exam AMA-DS, October 2023. Practice3"
author: "Ebner Lena"
date: "`r format(Sys.time(), '%d/%m/%Y')`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Life expectancy. Countries of the World

LE(age): Life expectancy at exact age \(x\) is defined as follows: The average number of remaining years of life expected by a hypothetical cohort of individuals alive at age \(x\) who would be subject during the remaining of their lives to the mortality rates of a given year. It is expressed as years.

We have downloaded life expectancy at exact age (for ages from 0 to 100 years) for different countries from the web page of United Nations, Department of Economic and Social Affairs, Extra Life Expectancy Division (2022). World Extra Life Expectancy Prospects 2022, Online Edition.

Other functions can be defined from LE(age):

Extra life expectancy: LE(age)-age.
Annual increase in LE: LE(age)-LE(age-1), for age\(\ge 1\).
Annual increase in LE, in logarithms: log( LE(age)-LE(age-1) ), for age\(\ge 1\).
As an example, here we have the life expectancy data of Spain for year 2021:

Here you have the **logarithm of annual increase in expected life length** for year 2021, as functions of age for all the countries in the data set

```{r}
 #load data
load("/home/le/Dev/advanced_multivariate_analysis/lab/partial_exam/ExamOctober2023.Rdata")
head(LE)
```

```{r}
matplot(1:100,t(lgDifLE), type="l",  col=8,
        xlab="Age",ylab="log-Increase in life expectancy",
        main="log-Increase in life expectancy, all countries, year 2021",
        cex.main=.75,cex.lab=.75,cex.axis=.75)
i <- which(rownames(lgDifLE)=='ES')
lines(1:100,lgDifLE[i,],lwd=4, col=1)
legend("bottomright",CtryName[i],lwd=4,col=1,bty="n")
```

The file `ExamOctober2023.Rdata` contains the following data:

- `LE`, a matrix (countries in rows, age in columns) with life expectancy at exact age
- `lgDifLE`, a matrix (countries in rows, age in columns) with logarithms of the annual increase in life expectancy.
- `CtryName`, a vector with the country names.
- `CtryISO2`, a vector with the 2-digits ISO country codes.

# Questions

## 1. Density estimation (2 points out of 10)

Consider the joint distribution of
`x=scale(LE[,1])`, `y=scale(lgDifLE[,15])`,
that is, the joint distribution of the centered and scaled values of the life expectancy at birth and the log-Increase in life expectancy at the age of 15.

Estimate the joint density of \((x,y)\) using a bivariate Gaussian kernel estimator with equal bandwidths in both dimensions, h=a\*c(1,1), with `a` chosen by maximum log-likelihood leave-one-out cross-validation in `seq(0.15,.3,by=0.025)`.

Which is the optimal value of `a`?
Give at least two graphical representations of the estimated density function. When it is possible, add the observed points to the graphics.

```{r}
x <- scale(LE[,1])
y <-scale(lgDifLE[,15])

X <- cbind(x,y)

library(sm)

plot(X,col=8)
sm.density(X,h=0.15*c(1,1),display="slice",props=c(25,50,75,95),col=2,add=TRUE)
```

```{r}
a_values <- seq(0.15,.3,by=0.025)
na <- length(a_values)
logLCVa <- numeric(length(a_values))

looCV_log__lik=list()

n <- dim(X)[1]

for (j in 1:na) {
  a <- a_values[j]
  for (i in 1:n){
    new.point <- matrix(X[i,],ncol=2)
    f.hat.i <- sm.density(X[-i,],h=a*c(1,1),display="none",eval.grid=FALSE,
                    eval.points=new.point)$estimate

     logLCVa[j] <- logLCVa[j] + log(f.hat.i)
  }
}

a_loo <- a_values[which.max(logLCVa)]
plot(a_values,logLCVa,type="b", main=paste("a_loo=",a_loo))
abline(v=a_loo,col=2)
```

```{r}
best=a_values[which.max(logLCVa)]
print(paste("best a (LOOCV)",best))
```

```{r}
plot(x,y,col="gray", asp=1)
# plot(X,col=8)
sm.density(X,h=best*c(1,1),display="slice",col=2,add=TRUE)
sm.density(X, h=best*c(1,1), display="slice", add=TRUE, col="red", props=95)
```

```{r}
sm.density(X, h=a_loo*c(1,1), display="persp")
```

## Fitting a GMM. (2 points out of 10)

Consider the previous bivariate data set \((x_i, y_i)\), \(i=1,\ldots,n\), where \(n\) is the number of countries.

- Fit a Gaussian Mixture Model (GMM) to these data choosing the best model by the Bayesian Information Criterion (BIC). Give graphical representations of the estimated model.

```{r}
library(mclust)
library(sm)
library(fpc)
library(ggplot2)
library(cluster)

rng=2:6
X <- cbind(x,y)
GMM <- Mclust(as.matrix(X), parameters=TRUE)

clust.ind <- GMM$classification
summary(GMM,parameters=TRUE)
```

```{r}
plot(GMM, what="BIC",asp=1)
```

```{r}
plot(GMM, what="classification",asp=1)
```

```{r}
plot(GMM, what="density",asp=1)
```

```{r}
plot(GMM, what="uncertainty",asp=1)
```

- How many mixture components have the fitted model?

```{r}
summary(GMM,parameters=TRUE)

# the fitted model has 2 mixture components
```

- Have the different components the same variance matrices?

```{r}
# The fitted GMM by Mclust is EVV (ellipsoidal, equal volume)
# indicates that the two variance matrices have equal volume (proportional of the product of their eigenvalues)
# varying shape (different the ratio of eigenvalues)
# varying orientation (eigenvectors in both matrices ar not parallel)
# # Therefore, the variance matrices are different.

# As seen above the the components have different variance matrices
# Variances:
# [,,1]
#            [,1]       [,2]
# [1,]  0.7467912 -0.5213116
# [2,] -0.5213116  0.4474711
# [,,2]
#            [,1]       [,2]
# [1,]  0.3729693 -0.2977819
# [2,] -0.2977819  0.5147658
```

- How the estimated density by GMM compares with the non-parametric estimator obtained before?

```{r}
par(mfrow = c(1, 2))

plot(GMM, what="density", sub="Mclust density estimation", asp=1, col=2)
points(x,y, col = "grey")

plot(X, sub="GMM density estimation", col = "grey")
sm.density(X,h=best*c(1,1),display="slice",props=c(50,75,95),col=2,add=TRUE)
```

```{r}
#
# The GMMM is unimodal, but the non-parametric density is 3-modal.
# Other characteristics are similar in both estimations.
```

- Could you give an interpretation of the clusters identified by the fitted GMM? Mentions some countries belonging to each cluster.

```{r}
plot(GMM, what="classification", asp=1, xlab="scale(LE[,1])",ylab="scale(lgDifLE[,15])")
points(X)
# text(X[,1], X[,2], rownames(X), pos = 3, cex = .7)

# there are 2 clusters
```

```{r}
#Some countries ate each cluster
# Blue cluster
Icl1 <- which(y<=-2)
cat("\n \n Red cluster\n")
CtryName[Icl1]
# Countries with high Life Expectancy at birth, larger than 82 years. So all the yearly increment of LE are small.
# For age=15 this increment are lower than 0.4 years for the selected countries.
```

```{r}
# Red cluster
Icl2 <- which(x<=-2)
cat("\n \n Blue cluster\n")
CtryName[Icl2]
# Countries with low Life Expectancy at birth, lower than 57 years.
# So the yearly increment of LE are large, in particular for age=15.
# this increment are larger than 3.7 years for the selected countries except one (which has 2.2 years).
```

## Db scan

- How many clusters are detected by Dbscan, when using epsilon=0.25 and minPts=4?

```{r}
library(fpc)
epsilon <- 0.25
minPts <- 4

dbscan_result <- fpc::dbscan(X, eps = epsilon, MinPts = minPts, showplot = 0)
plot(dbscan_result, X, main = sprintf("DBSCAN eps=%.2f minPts=%d", epsilon, minPts), frame = FALSE, xlab="x",ylab="y")

#dbscan.X <- dbscan::dbscan(X,eps = epsilon, minPts = minPts)
#plot(X, col=dbscan.X$cluster+1, pch=dbscan.X$cluster+1,
#     main=paste("dbscan::dbscan; epsilon=",epsilon,",minPts=",minPts),
#     xlab="x",ylab="y",asp=1)

# 4 clusters have been detected by dbcscan
# There are 4 clusters:
# length(unique(dbscan.X$cluster[dbscan.X$cluster>0]))
```

- How many outliers have been detected by Dbscan?

```{r}
dbscan_result

# table(dbscan.X$cluster)

# 22 outliers have been detected bz dbscan
sum(dbscan_result$cluster==0)
```

- Which is the relationship between the clusters identified by Dbscan and those identified by GMM?

```{r}
op<-par(mfrow=c(1,2))
plot(GMM, what="classification",asp=1, main="GMM",
     xlab="scale(LE[,1])",ylab="scale(lgDifLE[,15])")
plot(X, col=dbscan_result$cluster+1, pch=dbscan_result$cluster+1,
     main="dbscan",
     xlab="scale(LE[,1])",ylab="scale(lgDifLE[,15])",asp=1)
```

```{r}
par(op)

table(dbscan_result$cluster,GMM$classification)
# There is no a perfect coincidence between both clustering methods
```

## Dimensionality reduction (3 points out of 10)

Consider the data matrix lgDifLE. Do dimensionality reduction of the columns of lgDifLE, from 100 dimensions to \(q=1\), in two different ways:

**4.1** Using ISOMAP. Choose epsilon in seq(7, 12,by=1) as the value maximizing the correlation between the distance matrices in high and low dimensional spaces.

```{r}
Dlg <- dist(lgDifLE)

q<-1

library(vegan)
epsilon <- seq(7, 12,by=1)

corr.dists <- array(0,dim=length(epsilon))
isomap.eps <- array(vector("list",1),dim=dim(corr.dists))

for (i in 1:length(epsilon)){
  isomap.eps[[i]] <- isomap(as.matrix(Dlg), ndim=q, epsilon=epsilon[i])
  D2.eps <- dist(isomap.eps[[i]]$points)
  corr.dists[i] <- cor(Dlg,D2.eps)^2
  #print(c(i,j,LC[i,j]))
}

i.max <- which.max(corr.dists)
eps.max <- epsilon[i.max]
isomap.max <- isomap.eps[[i.max]]
isomap.corr.dists.max <- corr.dists[i.max]

plot(epsilon,corr.dists, type="b")
abline(v=eps.max, col=2)
```

```{r, warning=FALSE}
print(paste0("ISOMAP: epsilon.max=",eps.max,", corr.dists(epsilon.max)=",corr.dists[i.max]))
```

```{r}
pairs(cbind(X[,1:2] ,isomap=isomap.max$points[,1]), pch=20, main="Best ISOMAP output in 1-dim")
```

```{r}
plot(isomap.max$points,isomap.max$points)
text(isomap.max$points,isomap.max$points,CtryName,
     pos=3-sign(isomap.max$points), cex=.75)
```

Select 3 countries located in the low dimensional space in such a way that they cover the variability of all the points in this dimension. Then plot their lgDifLE functions in a graphic as that in page 2.

```{r}
# Select 3 countries located in the low dimensional space in such a way that they cover the variability of all the points in this dimension.
imin <- which.min(isomap.max$points)
imed <- which.min(abs(isomap.max$points -
                        median(isomap.max$points)))
imax <- which.max(isomap.max$points)
i3 <- c(imin,imed,imax)
matplot(1:100,t(lgDifLE), type="l",  col=8,
        xlab="Age",ylab="log-Increase in life expectancy",
        main="log-Increase in life expectancy, all countries, year 2021",
        cex.main=.75,cex.lab=.75,cex.axis=.75)
for (i in 1:3) lines(1:100,lgDifLE[i3[i],],lwd=4, col=i)
legend("bottomright",CtryName[i3],lwd=4,col=1:3,bty="n")
```

How do you interpret the dimension obtained with ISOMAP?

```{r}
# The one-dimentional scores obtained by isomap sorts the countries from those having lower LE(age) and higher lgDifLE(age), to those having higher LE(age) and lower lgDifLE(age)
```

**4.2** USING t-SNE. Use theta=0 in Rtsne function and choose perplexity in seq(10,50,by=10) as the value maximizing the correlation between the distance matrices in high and low dimensioal spaces.

```{r}
library("Rtsne")
D=dist(as.matrix(X))

set.seed(42)
theta= 0.0
perplexity <- seq(10,50,by=10)
q=1

corr.dists <- array(0,dim=length(perplexity))
tSNE.perp <- array(vector("list",1),dim=dim(corr.dists))

for (i in 1:length(perplexity)){
  tSNE.perp[[i]] <- Rtsne(D, dims=q,
                   perplexity=perplexity[i],
                   theta=theta, num_threads = 1)
    D2.perp <- dist(tSNE.perp[[i]]$Y)
    corr.dists[i] <- cor(Dlg,D2.perp)^2
    #print(c(i,j,LC[i,j]))
}

i.max <- which.max(corr.dists)
perp.max <- perplexity[i.max]
tSNE.max <- tSNE.perp[[i.max]]
tSNE.corr.dists.max <- corr.dists[i.max]

plot(perplexity,corr.dists, type="b")
abline(v=perp.max, col=2)
```

```{r}
print(paste0("tSNE: perplexity.max=",perp.max,", corr.dists(perplexity.max)=",corr.dists[i.max]))
```

```{r}
plot(tSNE.max$Y,tSNE.max$Y, main="t-SNE")
text(tSNE.max$Y,tSNE.max$Y,CtryName, #rownames(X)
     pos=3-sign(tSNE.max$Y), cex=.75)
```

```{r}
pairs(cbind(x=X[,1], y=X[,2] ,tsne=tSNE.max$Y[,1]), pch=20, main="Best t-SNE output in 1-dim")
```

- Select 3 countries located in the low dimensional space in such a way that they cover the variability of all the points in this dimension. Then plot their lgDifLE functions in a graphic as that in page 2.

```{r}
# Select 3 countries located in the low dimensional space in such a way that they cover the variability of all the points in this dimension.
imin <- which.min(tSNE.max$Y)
imed <- which.min(abs(tSNE.max$Y -
                        median(tSNE.max$Y)))
imax <- which.max(tSNE.max$Y)
i3 <- c(imin,imed,imax)
matplot(1:100,t(lgDifLE), type="l",  col=8,
        xlab="Age",ylab="log-Increase in life expectancy",
        main="log-Increase in life expectancy, all countries, year 2021",
        cex.main=.75,cex.lab=.75,cex.axis=.75)
for (i in 1:3) lines(1:100,lgDifLE[i3[i],],lwd=4, col=i)
legend("bottomright",CtryName[i3],lwd=4,col=1:3,bty="n")
```

- How do you interpret the dimension obtained with t-SNE?

```{r}
# The one-dimentional scores obtained by tSNE sorts the countries from those having lower LE(age) and higher lgDifLE(age), to those having higher LE(age) and lower lgDifLE(age)
```

**4.3** Do a pairs plot (that is, a matrix of scatterplots) of the matrix having the following 4 columns:

- LE[,1], that is, life expectancy at birth.
- First principal component of lgDifLE.
- 1-dimensional configuration obtained by isomap.
- 1-dimensional configuration obtained by t-SNE.
  Comment on the resulting graphic.

```{r}
# compare to PCA
# pca_le = princomp(lgDifLE)$scores[,1]
# plot(pca_le, pca_le, t='n')
# text(pca_le,pca_le, labels=rownames(X))

aux <- cbind(LE[,1], princomp(lgDifLE)$scores[,1],isomap.max$points,tSNE.max$Y)
colnames(aux) <- c("LE_birth", "PC.1", "Isomap", "tSNE")
pairs(aux)
```

```{r}
# The one-dimentional scores obtained by the three methods (PC1, isomap and tSNE) sorts the countries approximately in the same order as LE at birth.
# PC1 is almost equialt to isomap.
```
