---
title: "Final Exam AMA-DS, January 2023. Practice"
author: "Pedro Delicado"
output:
  html_document:
    number_sections: yes
  html_notebook: null
  pdf_document:
    fig_caption: yes
    number_sections: yes
classoption: a4paper
---

# Human Development Index

In the HDI.Rdata file you will find some data published by the UNITED NATIONS DEVELOPMENT PROGRAM in its Human Development Report corresponding to the year 2017. After reading this file with

```{r}
load("HDI.Rdata")
```

you will find the data frame `HDI` with the following the following variables:

  
| Variable | Description |
| --- | --- |
| HDI | Human Development Index (HDI) |
| Median_age | Median age (years) |
| Old\_age\_dep_rat | Old-age (65 and older) dependency ratio (per 100 people ages 15-64) |
| Exp\_years\_school | Expected years of schooling (years) |
| Life_expec | Life expectancy at birth (years) |
| Inf\_Mort\_rat | Mortality rate, infant (per 1,000 live births) |
| log\_GDP\_per_cap | logarithm of the Gross domestic product (GDP) per capita (PPP $: international dollars with purchasing power parity) |
| Mob\_phon\_100_peopl | Mobile phone subscriptions (per 100 people) |
| Agric\_employ\_perc | Employment in agriculture (% of total employment) |
| Servi\_employ\_perc | Employment in services (% of total employment) |
| Employment_rat | Employment to population ratio (% ages 15 and older) |
| Continent | Continent |

The country names are the row names of `HDI` and can be recovered by

```{r}
country <- row.names(HDI)
```

## 1. Dimensionality reduction (2 points out of 10)

Let X be the data matrix with columns 2 to 11 of `HDI`:

```{r}
X <- as.matrix(HDI[,2:11])
```
Do dimensionality reduction of the columns of XX, from 10 dimesnions to q=1q=1, in two different ways:

**1.1** Computing the principal curve (as defined by Hastie and Stuetzle, 1989). Select the tuning parameter by using the local continuity meta criteria (with K′=5K′=5). Choose `df` in `seq(15,25,by=2)`

```{r}
library(princurve)
library(stops)
```

```{r}
LCMC <- function(D1,D2,Kp=10){
  D1 <- as.matrix(D1)
  D2 <- as.matrix(D2)
  n <- dim(D1)[1]
  N.Kp.i <- numeric(n)
  for (i in 1:n){
    N1.i <- sort.int(D1[i,],index.return = TRUE)$ix[1:Kp]
    N2.i <- sort.int(D2[i,],index.return = TRUE)$ix[1:Kp]
    N.Kp.i[i] <- length(intersect(N1.i, N2.i))
  }
  N.Kp<-mean(N.Kp.i)
  M.Kp.adj <- N.Kp/Kp - Kp/(n-1)
  
  return(list(N.Kp.i=N.Kp.i, M.Kp.adj=M.Kp.adj))
}
```

```{r}
# principal curve
Kp <- 5
DX <- dist(X)

vdf <- seq(15,25,by=2)
LC.prcv <- numeric(length(vdf))

prcv.df <- vector("list",length(LC.prcv))

for (j in 1:length(vdf)){
    #prcv.df[[j]] <- principal_curve(prcomp(HDI[, 2:11])$x, df=vdf[j])
    prcv.df[[j]] <- principal_curve(X, df=vdf[j])
    D2.j <- dist(prcv.df[[j]]$lambda)
    LC.prcv[j] <- LCMC(DX,D2.j,Kp)$M.Kp.adj
}

plot(vdf,LC.prcv,type="b")
```

```{r}
j.max <- which.max(LC.prcv)
df.max <- vdf[j.max] 
PrCv.max <- prcv.df[[j.max]]
print(paste0("df.max=",df.max))
```

**1.2** Performing local MDS. Select the tuning parameters by using the local continuity meta criteria (with K′=5K′=5). Choose `K` in `seq(3,15,by=2)` and `tau` in `c(.5,1)`.

```{r}
# local MDS
Kp <- 5
# DX <- dist(X)

K <- seq(3,15,by=2)#c(5,10,15)
tau <- c(.5,1)#c(.1,.5,1)
q <- 1
conf0 <- cmdscale(DX, k=q)$points

LC <- matrix(0,nrow=length(K),ncol=length(tau))
lmds.k.tau <- array(vector("list",1),dim=dim(LC))

init.conf <- conf0

for (i in 1:length(K)){
  for (j in 1:length(tau)){
    lmds.k.tau[[i,j]] <- lmds(as.matrix(DX), init=conf0, ndim=q, 
                              k=K[i], tau=tau[j], itmax = 100)
    D2.k.tau <- lmds.k.tau[[i,j]]$confdist
    LC[i,j] <- LCMC(DX,D2.k.tau,Kp)$M.Kp.adj
    #print(c(i,j,LC[i,j]))
  }
}

ij.max <- arrayInd(which.max(LC),.dim=dim(LC))
k.max <- K[ij.max[1]] 
tau.max <- tau[ij.max[2]] 
lmds.max <- lmds.k.tau[[ij.max[1],ij.max[2]]]
lmds.LC.max <- LC[ij.max]

print(paste0("Local MDS: k.max=",k.max,", tau.max=",tau.max, ",LC(k.max,tau.max)=",LC[ij.max]))
```

**1.3** Do a pairs plot (that is, a matrix of scatterplots) of the matrix having the following 4 columns:
- `HDI$HDI`
- First principal component of XX
- 1-dimensional configuration obtained by principal curves.
- 1-dimensional configuration obtained by local MDS.

```{r}
aux <- cbind(HDI$HDI, prcomp(X)$x[,1],PrCv.max$lambda,lmds.max$conf)
colnames(aux) <- c("HDI", "PC.1", "PrCv", "lmds.1")
pairs(aux)
```

**1.4** Comment your results

```{r}
# The HDI is defined in such a way it represents a summary of all the 
# measurable variables indicating "human development", in some way.
# This objective can also be reached by using any dimensionality reduction
# method, as PC, principal curves or local MDS.
# Then it is natural that these 4 variables are highly correlated.
```

## 2. Nonparametric regression (2.5 points out of 10)

**2.1** Do the scatterplot of y=y=`Old_age_dep_rat` against x=x=`Median_age`. Do yout think there are some countries which could be considered outliers?
```{r}
plot(HDI$Median_age,HDI$Old_age_dep_rat)
```

```{r}
plot(HDI$Median_age,HDI$Old_age_dep_rat)
text(HDI$Median_age,HDI$Old_age_dep_rat,country)
```

**2.2** To identify those possible outliers, fit a simple nonparametric regression model of `Old_age_dep_rat` as a function of `Median_agey`. Use smoothing splines with degrees of freedom chosen by Generalized Cross Validation.

```{r}
library(mgcv)
```

```{r}
gam1 <- gam(Old_age_dep_rat~s(Median_age),data=HDI)
summary(gam1)
```

```{r}
plot(gam1, residuals=TRUE, cex=4)
```
```{r}
# another alternative: using smooth.spline()
sm.spl.1 <- smooth.spline(HDI$Median_age,HDI$Old_age_dep_rat)
```

**2.3** Once you have fitted this model, which is the estimated value σ^σ^ of the residual standard deviation?
If this estimation is not included in the fitted smoothing spline object, you can compute it from the residuals. If the residuals are not directly available, you can use the functions `fitted` to compute the fitted values of your estimated model and use them to compute residuals.

```{r}
(sig <- sqrt(gam1$sig2))
```
```{r}
# another alternative: using smooth.spline()
n<- dim(HDI)[1]
fitted_values <- fitted(sm.spl.1)
resids <- HDI$Old_age_dep_rat - fitted_values
(sig_spl <- sqrt(sum(resids^2)/(n-sm.spl.1$df)))
```

**2.4** Add to the scatterplot of y=y=`Old_age_dep_rat` against x=x=`Median_age` the following elements:
- the points with coordinates x=x=`Median_age` and y=y=the fitted values.
- the points with coordinates x=x=`Median_age` and y=y= the fitted values +/−1.96σ^+/−1.96σ^.
- the names of the countries for which the absolute value of the residuals are larger than 1.96 σ^1.96 σ^. These points are those you can consider outliers.

```{r}
plot(HDI$Median_age,HDI$Old_age_dep_rat)
points(HDI$Median_age,gam1$fitted.values,col=2, pch=19)
points(HDI$Median_age,gam1$fitted.values-1.96*sig,col=4, pch="+")
points(HDI$Median_age,gam1$fitted.values+1.96*sig,col=4, pch="+")
I_res_neg <- which(gam1$fitted.values-1.96*sig>HDI$Old_age_dep_rat)
I_res_pos <- which(gam1$fitted.values+1.96*sig<HDI$Old_age_dep_rat)
text(HDI$Median_age[I_res_neg],HDI$Old_age_dep_rat[I_res_neg],
     country[I_res_neg], pos=4)
text(HDI$Median_age[I_res_pos],HDI$Old_age_dep_rat[I_res_pos],
     country[I_res_pos], pos=2)
```
```{r}
country[I_res_neg]
```


```{r}
country[I_res_pos]
```


```{r}
# another alternative: using smooth.spline()
plot(HDI$Median_age,HDI$Old_age_dep_rat)
points(HDI$Median_age,fitted_values,col=2, pch=19)
points(HDI$Median_age,fitted_values-1.96*sig_spl,col=4, pch="+")
points(HDI$Median_age,fitted_values+1.96*sig_spl,col=4, pch="+")
I_res_neg <- which(fitted_values-1.96*sig_spl>HDI$Old_age_dep_rat)
I_res_pos <- which(fitted_values+1.96*sig_spl<HDI$Old_age_dep_rat)
text(HDI$Median_age[I_res_neg],HDI$Old_age_dep_rat[I_res_neg],
     country[I_res_neg], pos=4)
text(HDI$Median_age[I_res_pos],HDI$Old_age_dep_rat[I_res_pos],
     country[I_res_pos], pos=2)
```
```{r}
country[I_res_neg]
```

```{r}
country[I_res_pos]
```

**2.5** Could you provide any explanation why those countries are outliers?

```{r}
# There are 7 countries with very low "Old age dependence ratio":
## [1] "United Arab Emirates" "Bahrain"              "Korea (Republic of)" 
## [4] "Kuwait"               "Oman"                 "Qatar"               
## [7] "Singapore"
#
# Five of them are in the Persian Gulf. They are those with lowest median age (around 30 years).
# They are young countries in median with very few old people. 
# Probably foreign immigration could explain these facts.
# The other two countries are well developed Asian countries: South Korea and Singapore. 
#
# There are 4 countries with unusually high "Old age dependence ratio":
## [1] "Argentina" "Israel"    "Japan"     "Uruguay"
#
# Japan is the most aged country in the data set: a median age over 45 years. 
# The other three countries have median age in 30-35 years.
```

## 3. Functional data analysis (2.5 points out of 10)

The Rdata file `HDI_series.Rdata` contains the data frame `HDI_series`, with rows corresponding to the 134 countries (the row names of `HDI_series` are the country names). The 32 columns of `HDI_series` correspond to years 1990 to 2021. The entry (i,j)(i,j) of `HDI_series` is the value of HDI for country ii at year jj.
(Note: The source of data of `HDI` and `HDI_series` are different. Then the values of `HDI_series` for year 2017 do not coincide exactly with those of `HDI$HDI`)

**3.1** Transform the data in `HDI_series` into a functional data set, with countries as individuals and years as argument of the functions:

- First, transform the raw data to a `fdata`object.
- Then, smooth the raw functional data in an optimal way using generalized cross-validation and local linear smoothing.

Call `HDI_LL` to the smoothed functional data.

```{r}
load("HDI_series.Rdata")
library(fda.usc)
years <- 1990:2021
HDI.fdata <- fdata(mdata=as.matrix(HDI_series), argvals = years, 
                      rangeval = range(years), 
                      names = list(main="HDI 1990-2021",
                                   xlab="Year",ylab="HDI"))

HDI.LLR <- optim.np(HDI.fdata, type.S = S.LLR)

plot(HDI.LLR$h,HDI.LLR$gcv,type="b")
abline(v=HDI.LLR$h.opt,col=2)
text(HDI.LLR$h.opt,min(HDI.LLR$gcv),
      paste("h.opt=",round(HDI.LLR$h.opt,2)),
      pos=3,col=2)
```

```{r}
HDI_LL <- HDI.LLR$fdata.est
plot(HDI_LL)
```

**3.2** Do descriptive statistics of `HDI_LL` (mean, median and standard deviation functions).
(Note: If you have not found `HDI_LL` then use the raw data functional data set)

```{r}
plot(HDI_LL, col=8)
plot(func.mean(HDI_LL),add=TRUE,lwd=4,col=2)
plot(func.med.FM(HDI_LL),add=TRUE,lwd=4,col=4)
legend("bottomright",c("mean","median FM"),lwd=4,col=c(2,4),cex=1)
```

```{r}
plot(sqrt(func.var(HDI_LL)))
```

```{r}
plot(HDI_LL, col=8)
plot(func.mean(HDI_LL),add=TRUE,lwd=4,col=2)
plot(func.mean(HDI_LL)+1*sqrt(func.var(HDI_LL)),add=TRUE,lwd=4,col=6,lty=2)
plot(func.mean(HDI_LL)-1*sqrt(func.var(HDI_LL)),add=TRUE,lwd=4,col=6,lty=2)
legend("bottomright",c("mean","mean +/- stdev"),lwd=4,col=c(2,6), lty=c(1,2),cex=1)
```

**3.3** Look for outliers in the functional data set `HDI_LL`.
(Note: If you have not found `HDI_LL` then use the raw data functional data set)

```{r}
outl_HDI_LL.mode <- outliers.depth.pond(HDI_LL, nb = 200, dfunc = depth.mode)
outl_HDI_LL.mode$outliers
```

```{r}
# No outliers found!
```

```{r}
outl_HDI_LL.FM <- outliers.depth.pond(HDI_LL, nb = 200, dfunc = depth.FM)
outl_HDI_LL.FM$outliers
```

```{r}
# No outliers found!
```

**3.4** Perform Functional PCA on `HDI_LL` and try to give an interpretation of the first two principal functions.
(Note: If you have not found `HDI_LL` then use the raw data functional data set)

```{r}
fd_HDI_LL <- fdata2fd(HDI_LL)
fpcaHDIfd <- pca.fd(fd_HDI_LL)
par(mfrow=c(1,2))
plot(fpcaHDIfd)
```

```{r}
par(mfrow=c(1,1))

fpcaHDI <- fdata2pc(HDI_LL)
summary(fpcaHDI)
```

**3.5** Do the scatterplot of the scores on the first two principal functions, adding the country name to each point. Comment your results.

```{r}
plot(fpcaHDIfd$scores)
text(fpcaHDIfd$scores[,1], fpcaHDIfd$scores[,2], 
     rownames(HDI_series))
```

```{r}
plot(fpcaHDI$x)
text(fpcaHDI$x[,1], fpcaHDI$x[,2], 
     rownames(HDI_series))
```

```{r}
# 1st FPC: Index of Human Development:
plot(fpcaHDI$x[,1],HDI_series[,32])
```
```{r}
#  Less developed countries: Niger, Mozambique, Burundi, Central African Republic, ...
#  Most developed countries: Norway, Switzerland, Sweden, Australia, USA, ...
# 2nd FPC: Speed of development growth: 
# Most slow development: Leshoto, Libia, Eswatini, Namibia, Central African Republic, ...   
# Most fast development: China, Turkey, Myanmar, Singapore, ...
```

## 4. Interpretable Machine Learning (2 points out of 10)

Consider the problem of predicting `HDI$HDI` from the rest of variables in the data frame `HDI`.

**4.1** Found the gam model you consider the best one to explain `HDI$HDI` from the rest of variables in the data frame `HDI`. Call if `gamHDI`.

```{r}
gam1 <- gam(HDI~
              s(Median_age) + 
              s(Old_age_dep_rat) + 
              s(Exp_years_school) + 
              s(Life_expec) + 
              s(Inf_Mort_rat) + 
              s(log_GDP_per_cap,by=Continent) + 
              s(Mob_phon_100_peopl) + 
              s(Agric_employ_perc) + 
              s(Servi_employ_perc) + 
              s(Employment_rat) + 
              Continent,
            data=HDI)
summary(gam1)
```

```{r}
gam2 <- gam(HDI~
              s(Old_age_dep_rat) + 
              s(Exp_years_school) + 
              Life_expec + 
              s(log_GDP_per_cap,by=Continent) + 
              s(Mob_phon_100_peopl) + 
              Agric_employ_perc*Continent + 
              s(Employment_rat,by=Continent) + 
              Continent,
            data=HDI)
summary(gam2)
```
```{r}
plot(gam2)
```

```{r}
gam3 <- gam(HDI~
              s(Old_age_dep_rat) + 
              s(Exp_years_school) + 
              Life_expec + 
              s(log_GDP_per_cap) + 
              Agric_employ_perc*Continent + 
              s(Employment_rat,by=Continent) + 
              Continent,
            data=HDI)
summary(gam3)
```
```{r}
anova(gam3,gam2,test="F")
```

```{r}
gam4 <- gam(HDI~
              s(Old_age_dep_rat) + 
              s(Exp_years_school) + 
              Life_expec + 
              s(log_GDP_per_cap) + 
              Continent,
            data=HDI)
summary(gam4)
```

```{r}
gamHDI <- gam4
```

**4.2** Fit a random forest to explain `HDI$HDI` from the rest of variables in the data frame `HDI`. Call if `rfHDI`.

```{r}
library(ranger)
library(randomForest)
library(caret)
library(vip)
library(DALEX)
library(DALEXtra)
library(lime)
library(iml)
library(localModel)
# library(fastshap) # Attention! It re-define "explain" from DALEX
```

```{r}
rfHDI <- ranger(
  HDI ~ .,
  data = HDI, 
  importance='impurity'
)
print(rfHDI)
```

**4.3** Compute the relevance of each explanatory variable by Shapley values at both models, `gamHDI` and `rfHDI`. Comment your results.

```{r}
rf_imp <- vip(rfHDI, num_features = 11)
rf_shapley <- vip(rfHDI, method="shap",
                  pred_wrapper=yhat, num_features = 11,
                  newdata=HDI[,-1]) # Package "fastshap" needed for this function to work
gam_shapley <- vip(gamHDI, method="shap",
                  pred_wrapper=predict.gam, 
                  num_features = 11,
                  newdata=HDI[,-1],
                  exact=TRUE) # Package "fastshap" needed for this function to work

grid.arrange(rf_imp, rf_shapley, gam_shapley,
             ncol=2, nrow=2,
             top="Top left: RF Impurity. Top right: RF Shapley values. Bottom left: GAM Shapley values"
            )
```

**4.4** Consider these two countries: Spain and Qatar. For each of them, provide local explanations for the predicted values when using the random forest `rfHDI` (choose the explanatory methods that you consider most appropriate.) Comment your results.

```{r}
# Using library DALEX
# help(package="DALEX")
# ? explain
explainer_rf <- explain.default(model = rfHDI, data = HDI[, -1], y = HDI[,1], 
                               label = "Random Forest")
```

```{r}
aux <- which(country=="Spain")
(Spain <- HDI[aux,])
```

```{r}
aux <- which(country=="Qatar")
(Qatar <- HDI[aux,])
```

```{r}
## SHAP 
bd_rf <- predict_parts(explainer = explainer_rf,
                 new_observation = Spain,
                            type = "shap")

bd_rf
```
```{r}
plot(bd_rf)
```
```{r}
bd_rf <- predict_parts(explainer = explainer_rf,
                 new_observation = Qatar,
                            type = "shap")

bd_rf
```
```{r}
plot(bd_rf)
```

```{r}
## Break-down plots
bd_rf <- predict_parts(explainer = explainer_rf,
                 new_observation = Spain,
                            type = "break_down")

bd_rf
```

```{r}
plot(bd_rf)
```

```{r}
bd_rf <- predict_parts(explainer = explainer_rf,
                 new_observation = Qatar,
                            type = "break_down")

bd_rf
```

```{r}
plot(bd_rf)
```

## LIME

```{r}
lime_rf <- predict_surrogate(explainer = explainer_rf, 
                  new_observation = Spain[,-1], 
                  type = "localModel")
                  #type = "iml") # it does not work
                  #n_features = 6, 
                  #n_permutations = 1000,
                  #type = "lime") # it does not work

lime_rf 
```

```{r}
plot(lime_rf)
```

```{r}
lime_rf <- predict_surrogate(explainer = explainer_rf, 
                  new_observation = Qatar[,-1], 
                  type = "localModel")
                  #type = "iml") # it does not work
                  #n_features = 6, 
                  #n_permutations = 1000,
                  #type = "lime") # it does not work

lime_rf
```

```{r}
plot(lime_rf)
```

```{r}
## Local Graphics
### Individual conditional expectation (ICE) plot, or ceteris paribus plot

cp_rf <- predict_profile(explainer = explainer_rf, 
                           new_observation = Spain)
cp_rf
```

```{r}
plot(cp_rf)
```


```{r}
cp_rf <- predict_profile(explainer = explainer_rf, 
                           new_observation = Qatar)
cp_rf
```


```{r}
plot(cp_rf)
```

