---
title: "Splines 1: Spline Interpolation"
author: "Pedro Delicado"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
  pdf_document:
    fig_caption: yes
    number_sections: yes
classoption: a4paper
---

<!-- Comment lines are like this one -->
<!-- Use "\newpage" when you want a new page break in the pdf output  -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Splines in R: Libraries and functions

The following R libraries and functions implement the techniques related with splines.

1. Library `stats`. Installed by default in R and charged when starting R. It includes the folowing functions, among other:

- `spline`: Return the cubic interpolating spline for a data set.
- `smooth.spline`: Fit a cubic spline to data set solving the penalized least squares problem.
- `predict.smooth.spline`: Predict the value of a spline function fitted by `smooth.spline` for new values of the explanatory variable.

2. Library `splines`. See `help(package=splines)` for a complete list of functions in the library. Among them we remark the following:

- `bs`, `ns`: Build bases of B-splines and natural B-splines.
  (The example of the combined use of `bs` and `lm` to fit a smoothing spline to a data set has special interest.)
- `interpSpline`: Return the cubic interpolating spline for a data set.

# Interpolation by splines

Example of interpolating natural cubic spline:

```{r}
# help(spline)
# set.seed(3333) # nice results with this seed
n <- 9
x <- runif(n)
y <- rnorm(n)
spl.xy <- spline(x, y, method ="natural",
                 xmin=0, xmax=1, n = 201)
plot(x, y, pch=19, xlim=c(0,1), ylim=range(spl.xy$y),
     main = paste("Natural spline interpolation through", n, "points"))
lines(spl.xy, col = 2)
```

Natural spline extrapolation is linear:

```{r}
spl.xy <- spline(x, y, method ="natural",
                 xmin=-1, xmax=2, n = 601)
plot(x, y, pch=19, xlim=c(-1,2), ylim=range(spl.xy$y),
     main = "Natural spline extrapolation is linear")
lines(spl.xy, col = 2)
```
