---
title: 'Splines 6: Spline smoothing for glm. Country Development Data'
author: "Pedro Delicado"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
  pdf_document:
    fig_caption: yes
    number_sections: yes
classoption: a4paper
---
<!-- Comment lines are like this one -->
<!-- Use "\newpage" when you want a new page break in the pdf output  -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Development data in 179 countries

We will be working with the file `HDI.2017.subset.csv` containing information on development indicators measured in 179 countries (Source: [Human Development Data (1990-2017)](http://hdr.undp.org/en/data), The Human Development Report Office, United Nations).

This file contains the following variables:

* `Life.expec` Life expectancy at birth.
* `Life.expec.f` Life expectancy at birth for females.
* `Life.expec.m` Life expectancy at birth for males.
* `le.fm` Difference `Life.expec.f` minus `Life.expec.m`.
* `Inf.Mort.rat` Infant mortality rate: The annual number of deaths of infants under one year of age per 1,000 live births in the same year.
* `Agric.employ.%` Employment in agriculture (% of total employment).

```{r}
countries<-read.csv2(file="HDI.2017.subset.csv",row.names = 1)
attach(countries)
head(countries)
names(countries)
# [1] "country_name"   "Life.expec"     "Life.expec.f"  
# [4] "Life.expec.m"   "le.fm"          "Inf.Mort.rat"  
# [7] "Agric.employ.."     
```


# Local binary regression using splines. 

## Classify countries according to `Agric.employ..`. 

The median value of variable `Agric.employ..` is `r median(Agric.employ..)`. 
Create a new binary variable `ind.agr` indicating for any country if variable `Agric.employ..` is lower than `r median(Agric.employ..)` (1) or not (0).
This new variable is a development indicator.

```{r}
ind.agr<-(Agric.employ..<median(Agric.employ..))
```

## Classify countries according to `Life.expec`.

The median value of variable `Life.expec` is `r median(Life.expec)`. 
Create a new binary variable `ind.le` indicating for any country if variable `Life.expec` is greater than `r median(Life.expec)` (1) or not (0).
This new variable is a development indicator.

```{r}
ind.esp<-(Life.expec> median(Life.expec) )
```

## Local binary regression using splines.

Combine functions `bs` and `glm` to fit the following local binary regression models:

* `ind.agr` as a function of `Life.expec`,
* `ind.le`  as a function of `Agric.employ..`.


### Fit a generalized nonparametric regression model to explain `ind.agr` as a function of `Life.expec`.

First, We sort the data according to the explanatory variable.
```{r} 
x <- Life.expec
y <- ind.agr
sx <- sort(x,index.return =TRUE)
x <- sx$x
y <- y[sx$ix]
```

**First approach:** Combining functions `bs` and `glm`.

In this approach the degree of smoothing is controlled by the number of knots. 

```{r} 
library(splines)
# k=3: Cubic splines
k <- 3
my.knots <- unique(x)
my.knots <- my.knots[-c(1,length(my.knots))] 
basis <- bs(x=x,knots=my.knots,intercept=T,degree=k)
glm.spl.1 <- glm(y~basis)
```

**Second aproach:** Spline smoothing fit of a generalized nonparametric regression.

```{r,warning=FALSE}
source('IRWLS_logistic_regression.R')
my.spline.glm <- 
  logistic.IRWLS.splines(x,y,
                         df=min(length(unique(x)),
                                length(my.knots)+k+1),
                         all.knots=TRUE)
```

Plotting the estimations.
```{r}
#op <- par(mfrow=c(2,2))
plot(x,y,col=2,xlab="Life.expec",ylab="ind.agr")
lines(x,glm.spl.1$fitted.values)
abline(v=my.knots,lty=2,col="grey")
lines(x,my.spline.glm$fitted.values,col=4)
```

### Repeat the previous steps with lower number of knots.
```{r,message=FALSE,warning=FALSE,fig.asp=0.75}
op <- par(mfrow=c(2,2))
plot(x,y,col=2,xlab="Life.expec",ylab="ind.agr")
lines(x,glm.spl.1$fitted.values)
abline(v=my.knots,lty=2,col="grey")
lines(x,my.spline.glm$fitted.values,col=4)

#my.knots <- seq(min(x),max(x),length=length(x)/10)
my.knots <- quantile(x,seq(0,1,length=length(x)/10))
my.knots <- my.knots[-c(1,length(my.knots))]

basis <- bs(x=x,knots=my.knots,intercept=T,degree=k)
glm.spl.1 <- glm(y~basis,family="binomial")

plot(x,y,col=2,xlab="Life.expec",ylab="ind.agr")
lines(x,glm.spl.1$fitted.values)
abline(v=my.knots,lty=2,col="grey")

# Spline fit of gener.non.param. regression
my.spline.glm <- logistic.IRWLS.splines(x,y,df=min(length(unique(x)),length(my.knots)+k+1),all.knots=TRUE)
lines(x,my.spline.glm$fitted.values,col=4)

#my.knots <- seq(min(x),max(x),length=length(x)/20)
my.knots <- quantile(x,seq(0,1,length=length(x)/20))
my.knots <- my.knots[-c(1,length(my.knots))] 
basis <- bs(x=x,knots=my.knots,intercept=T,degree=k)
glm.spl.1 <- glm(y~basis,family="binomial")
plot(x,y,col=2,xlab="Life.expec",ylab="ind.agr")
lines(x,glm.spl.1$fitted.values)
abline(v=my.knots,lty=2,col="grey")
# Spline fit of gener.non.param. regression
my.spline.glm <- logistic.IRWLS.splines(x,y,df=min(length(unique(x)),length(my.knots)+k+1),all.knots=TRUE)
lines(x,my.spline.glm$fitted.values,col=4)

#my.knots <- seq(min(x),max(x),length=length(x)/40)
my.knots <- quantile(x,seq(0,1,length=length(x)/40))
my.knots <- my.knots[-c(1,length(my.knots))] 
basis <- bs(x=x,knots=my.knots,intercept=T,degree=k)
glm.spl.1 <- glm(y~basis,family="binomial")
plot(x,y,col=2,xlab="Life.expec",ylab="ind.agr")
lines(x,glm.spl.1$fitted.values)
abline(v=my.knots,lty=2,col="grey")
# Spline fit of gener.non.param. regression
my.spline.glm <- logistic.IRWLS.splines(x,y,df=min(length(unique(x)),length(my.knots)+k+1),all.knots=TRUE)
lines(x,my.spline.glm$fitted.values,col=4)

par(op)
``` 

### Fit a generalized nonparametric regression model to explain `ind.esp` as a function of `Agric.employ..`.

```{r,message=FALSE,warning=FALSE,fig.asp=0.75}
x <- Agric.employ..; 
y <- ind.esp; 
sx <- sort(x,index.return =TRUE)
x <- sx$x
y <- y[sx$ix]

op <- par(mfrow=c(2,2))

# k=3 
k <- 3
my.knots <- unique(x)
my.knots <- my.knots[-c(1,length(my.knots))] 
basis <- bs(x=x,knots=my.knots,intercept=T,degree=k)
glm.spl.1 <- glm(y~basis)
plot(x,y,col=2,xlab="Agric.employ..",ylab="ind.esp")
lines(x,glm.spl.1$fitted.values)
abline(v=my.knots,lty=2,col="grey")
# Spline fit of gener.non.param. regression
my.spline.glm <- logistic.IRWLS.splines(x,y,df=min(length(unique(x)),length(my.knots)+k+1),all.knots=TRUE)
lines(x,my.spline.glm$fitted.values,col=4)

#my.knots <- seq(min(x),max(x),length=length(x)/10)
my.knots <- quantile(x,seq(0,1,length=length(x)/10))
my.knots <- my.knots[-c(1,length(my.knots))] 
basis <- bs(x=x,knots=my.knots,intercept=T,degree=k)
glm.spl.1 <- glm(y~basis,family="binomial")
plot(x,y,col=2,xlab="Agric.employ..",ylab="ind.esp")
lines(x,glm.spl.1$fitted.values)
abline(v=my.knots,lty=2,col="grey")
# Spline fit of gener.non.param. regression
my.spline.glm <- logistic.IRWLS.splines(x,y,df=min(length(unique(x)),length(my.knots)+k+1),all.knots=TRUE)
lines(x,my.spline.glm$fitted.values,col=4)

#my.knots <- seq(min(x),max(x),length=length(x)/20)
my.knots <- quantile(x,seq(0,1,length=length(x)/20))
my.knots <- my.knots[-c(1,length(my.knots))] 
basis <- bs(x=x,knots=my.knots,intercept=T,degree=k)
glm.spl.1 <- glm(y~basis,family="binomial")
plot(x,y,col=2,xlab="Agric.employ..",ylab="ind.esp")
lines(x,glm.spl.1$fitted.values)
abline(v=my.knots,lty=2,col="grey")
# Spline fit of gener.non.param. regression
my.spline.glm <- logistic.IRWLS.splines(x,y,df=min(length(unique(x)),length(my.knots)+k+1),all.knots=TRUE)
lines(x,my.spline.glm$fitted.values,col=4)

#my.knots <- seq(min(x),max(x),length=length(x)/40)
my.knots <- quantile(x,seq(0,1,length=length(x)/40))
my.knots <- my.knots[-c(1,length(my.knots))] 
basis <- bs(x=x,knots=my.knots,intercept=T,degree=k)
glm.spl.1 <- glm(y~basis,family="binomial")
plot(x,y,col=2,xlab="Agric.employ..",ylab="ind.esp")
lines(x,glm.spl.1$fitted.values)
abline(v=my.knots,lty=2,col="grey")
# Spline fit of gener.non.param. regression
my.spline.glm <- logistic.IRWLS.splines(x,y,df=min(length(unique(x)),length(my.knots)+k+1),all.knots=TRUE)
lines(x,my.spline.glm$fitted.values,col=4)

par(op)
```