---
title: 'Splines 5: Penalized IRWLS algorithm for fitting a nonparametric version of logistic regression'
author: "Pedro Delicado"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
  pdf_document:
    fig_caption: yes
    number_sections: yes
subtitle: An Application to Burning Data
classoption: a4paper
---
<!-- Comment lines are like this one -->
<!-- Use "\newpage" when you want a new page break in the pdf output  -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Burn Injuries Data

The *Burn injuries dataset* (Fan, J. and I. Gijbels (1996). *Local polynomial modelling and its applications*) contains data from 435 adults (between ages 17 and 85) suffering from burn injuries
(you can find this dataset at file `burn.cont.dat`).
It includes two variables:

* `surv`: The binary response variable is taken to be 1 for those victims who survived their burn injuries and zero otherwise.
* `lgae`: log(area of third degree burn + 1) is taken as a covariate.

The conditional expectation of `surv` given a level of `lgae` is the conditional probability of 
	survival given this particular value of `lgae`.

```{r}
#
# reading the data
burns <-read.table("burn.cont.dat", head=T,sep=" ")
lgae <- burns[,2]
perm <- order(lgae)
lgae <- lgae[perm]
superv <- burns[perm,3]
n<-length(superv)
```

# Logistic regresion for Burn Injuries Data
## Logistic regresion using `glm`
```{r}
glm.burn <- glm(superv~lgae,family=binomial())
plot(lgae,superv,main="Regression 0-1 param. and non-param.")
lines(lgae,glm.burn$fitted.values,col=4)
```

## Logistic regression using our own IRWLS algorithm

The R script `"IRWLS_logistic_regression.R"` defines the function `logistic.IRWLS` that implements the IRWLS algorithm for fitting a logistic regression model by maximum likelihood. 

```{r}
# non parametric version
source("IRWLS_logistic_regression.R")
```

```{r}
x <- matrix(lgae,ncol=1)
y <- superv


my.glm.burn <- logistic.IRWLS(x,y)

plot(lgae,superv,main="Regression 0-1 param. and non-param.")
lines(lgae,glm.burn$fitted.values,col=4)
lines(lgae,my.glm.burn$fitted.values,col=2,lty=2)
```


# Spline fit of generalized non parametric regression

The R script `"IRWLS_logistic_regression.R"` also contains the definition of the function `logistic.IRWLS.splines` that implements the penalized IRWLS algorithm for fitting the non parametric version logistic regression model by local maximum likelihood.

See the complete `Description` of the function `logistic.IRWLS.splines` in the R script `"IRWLS_logistic_regression.R"`.

We fit now the non parametric regression model with binary response to the *Burn Injuries Data*.
First we use the default value for the *degrees of freedom* parameter, `df=6`, and we compare the fitted function with that obtained by parametric logistic regression.

```{r}
my.spline.glm.burn <- logistic.IRWLS.splines(x,y)

plot(lgae,superv,main="Regression 0-1 param. and non-param.")
lines(lgae,glm.burn$fitted.values,col=4)
lines(lgae,my.glm.burn$fitted.values,col=2,lty=2)
lines(lgae,my.spline.glm.burn$fitted.values,col=3) # non parametric function = green line
```

Now we call again function `logistic.IRWLS.splines` but now we fix a larger number of *degrees of freedom* (`df=20`) and ask the function to plot all the intermediate steps ('plts=TRUE`) in
both scales: the scale of $p(x)$ and the scale of $\theta(x)=\log(p(x)/(1-p(x)))$. 

```{r}
my.spline.glm.burn <- logistic.IRWLS.splines(x,y,df=20,plts = TRUE)

plot(lgae,superv,main="Regression 0-1 param. and non-param.")
lines(lgae,glm.burn$fitted.values,col=4)
lines(lgae,my.glm.burn$fitted.values,col=2,lty=2)
lines(lgae,my.spline.glm.burn$fitted.values,col=3)
```