---
title: "GAM for Boston housing data"
author: "Pedro Delicado"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
  pdf_document:
    fig_caption: yes
    number_sections: yes
classoption: a4paper
---
<!-- Comment lines are like this one -->
<!-- Use "\newpage" when you want a new page break in the pdf output  -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mgcv)
```

# Reading Boston Housing Data 

Reading the Corrected Boston Housing Data as they come in the library `mlbench` 
(see `library(mlbench); help(BostonHousing2)`; alternatively, see `library(spData); help(boston)`).

```{r}
# load("boston.Rdata")
library(mlbench)
# help(BostonHousing)
data(BostonHousing2)
boston.c <- BostonHousing2

names(boston.c)[12]<-'room'
names(boston.c)
# [1] "town"    "tract"   "lon"     "lat"     "medv"    "cmedv"   "crim"    "zn"     
# [9] "indus"   "chas"    "nox"     "room"    "age"     "dis"     "rad"     "tax"    
# [17] "ptratio" "b"       "lstat" 
```

We want to find a GAM model to explain `cmedv`, the *corrected median values of owner-occupied housing in USD 1000*, as a function of the other variables in the dataset. 

# Fitting a full GAM model for the Boston housing data

We start fitting a GAM model including all the explanatory variables 
(except `town`, `tract`, `medv`).

* Variable `chas`is a factor, so it has not to be smoothed.
* Variables `rad` has only 9 different values 
(`length(unique(boston.c$rad))=` `r length(unique(boston.c$rad))`). 
So we need to reduce the default value for the dimension `k` of the B-spline basis used to smooth this variable. 
* We fit jointly longitude and latitude by `te(lon,lat)` in order to capture the spatial effect.

```{r}
full.gam <- gam(cmedv ~ te(lon,lat) + s(crim) + s(zn) + 
                  s(indus) + chas + s(nox) + s(room) + 
                  s(age) + s(dis) + s(rad,k=5) +
                  s(tax) + s(ptratio) + s(lstat) + s(b),
                data=boston.c)
```

Now we examine the fitted `gam` object numerically and graphically.

## Model summary

```{r}
summary(full.gam)
```

According to `R-sq.(adj)` and `Deviance explained`, the quality of the fitted model is high. Nevertheless some refinements could be done:

1. Certain variables could be removed from the model because the corresponding `p-values` are large: 
`age`, `chas`, `zn` and `b`.
2. Other smoothed terms (those with `edf` very close to 1) can be replaced by linear terms: `rad`, `ptratio`.
 
## Visualization of individual effects of each explanatory variable:

```{r}
plot(full.gam, select = 1, residuals = TRUE, se=FALSE)
plot(full.gam, residuals = TRUE, shade=TRUE, seWithMean=TRUE, pages = 7)
```

## Visualization of the joint effects of a pair of variables:

```{r}
vis.gam(full.gam, view=c("lon","lat"), plot.type = "persp", theta=30, phi=30)
vis.gam(full.gam, view=c("lon","lat"), plot.type = "contour")
```

```{r}
vis.gam(full.gam, view=c("lstat","room"), plot.type = "persp", theta=30, phi=30)
vis.gam(full.gam, view=c("lstat","room"), plot.type = "contour")
```

## Some model diagnostics

The function `gam.check()` produces some diagnostic information about the fitting procedure and results. See the help of this function.

```{r}
gam.check(full.gam)
```

# A sequence of refined gam models

Now we remove non-significant variables from the model (`age`, `chas`, `zn` and `b`) one at a time, in decreasing order of p-values, until all variables are significant. 

```{r}
gam2 <- gam(cmedv ~ te(lon,lat) + s(crim) + s(zn) + 
                  s(indus) + chas + s(nox) + s(room) + 
                  s(dis) + s(rad,k=5) +
                  s(tax) + s(ptratio) + s(lstat) + s(b),
                data=boston.c)
summary(gam2)
```


```{r}
gam3 <- gam(cmedv ~ te(lon,lat) + s(crim) + s(zn) + 
                  s(indus) + s(nox) + s(room) + 
                  s(dis) + s(rad,k=5) +
                  s(tax) + s(ptratio) + s(lstat) + s(b),
                data=boston.c)
summary(gam3)
```


```{r}
gam4 <- gam(cmedv ~ te(lon,lat) + s(crim) + 
                  s(indus) + s(nox) + s(room) + 
                  s(dis) + s(rad,k=5) +
                  s(tax) + s(ptratio) + s(lstat) + s(b),
                data=boston.c)
summary(gam4)
```


```{r}
gam5 <- gam(cmedv ~ te(lon,lat) + s(crim) + 
                  s(indus) + s(nox) + s(room) + 
                  s(dis) + s(rad,k=5) +
                  s(tax) + s(ptratio) + s(lstat),
                data=boston.c)
summary(gam5)
```

Then we change smoothed by linear terms for other variables 
with `edf` very close to 1 (`rad` and `ptratio`), in increasing order of edf's.


There are two variables with edf equal to 1: `s(rad)` and `s(ptratio)`. First we change `s(rad)` to linear because it is closer to be non-significant.

```{r}
plot(gam5)
```


```{r}
gam6 <- gam(cmedv ~ te(lon,lat) + s(crim) + 
                  s(indus) + s(nox) + s(room) + 
                  s(dis) + rad +
                  s(tax) + s(ptratio) + s(lstat),
                data=boston.c)
summary(gam6)
```


```{r}
gam7 <- gam(cmedv ~ te(lon,lat) + s(crim) + 
                  s(indus) + s(nox) + s(room) + 
                  s(dis) + rad +
                  s(tax) + ptratio + s(lstat),
                data=boston.c)
summary(gam7)
```

The values of `R-sq.(adj)` and `Deviance explained` are very similar to those of the full model:

`R-sq.(adj) =  0.894   Deviance explained = 90.7%`

Now all variables are significant and the smooth terms have edf's larger than 1. 

We will check the last model to see if the smooth terms need larger number of knots. 


```{r}
gam.check(gam7)
```
We increase parameter `k` for `te(lon,lat)`, `s(indus)`, `s(nox)`, `s(dis)`: 

```{r}
gam8 <- gam(cmedv ~ te(lon,lat, k=10) + s(crim) + 
                  s(indus, k=20) + s(nox, k=20) + s(room) + 
                  s(dis, k=20) + rad +
                  s(tax) + ptratio + s(lstat),
                data=boston.c)
summary(gam8)
```
```{r}
gam9 <- gam(cmedv ~ te(lon,lat, k=10) + s(crim) + 
                  s(indus, k=20) + s(nox, k=20) + s(room) + 
                  s(dis, k=20) + rad +
                  tax + ptratio + s(lstat),
                data=boston.c)
summary(gam9)
```

```{r}
gam.check(gam9)
```

```{r}
plot(gam9)
```


Now we test the null hypothesis that states the final model is correct again the alternative that states that the full model is better.

In order to be able to test these hypotheses, we fit again the full model with the larger values of `k`. 

```{r}
full.gam.k <- gam(cmedv ~ te(lon,lat, k=10) + s(crim) + s(zn) + 
                  s(indus, k=20) + chas + s(nox, k=20) + s(room) + 
                  s(age) + s(dis, k=20) + s(rad,k=5) +
                  s(tax) + s(ptratio) + s(lstat) + s(b),
                data=boston.c)
summary(full.gam.k)
```

```{r}
anova(gam9,full.gam.k,test="F")
```

We do not reject `gam9` against `full.gam.k` at significance level 0.05.

Some graphics for better interpretation of the second fitted model:

```{r}
plot(gam9, select = 1, residuals = TRUE, se=FALSE)
plot(gam9, residuals = TRUE, shade=TRUE, seWithMean=TRUE, pages = 4)
```

```{r}
vis.gam(gam9, view=c("lon","lat"), plot.type = "persp", theta=30, phi=30)
vis.gam(gam9, view=c("lon","lat"), plot.type = "contour")
```

```{r}
vis.gam(gam9, view=c("lstat","room"), plot.type = "persp", theta=30, phi=30)
vis.gam(gam9, view=c("lstat","room"), plot.type = "contour")
```

