---
title: "My own local `glm`"
author: "Pedro Delicado"
date: "`r format(Sys.time(), '%d/%b/%Y')`"
output:
  html_document:
    number_sections: yes
  pdf_document:
    fig_caption: yes
    number_sections: yes
classoption: a4paper
---
<!-- Comment lines are like this one -->
<!-- Use "\newpage" when you want a new page break in the pdf output  -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Write your own local logistic regression function

Use the following R script to generate artificial data.

```{r}
# Plots in the slides have been done with this seed:
# set.seed(2)
set.seed(2)
beta0=0; beta1=1; k=4; s=.5; n<-400
x<-sort(rnorm(n))
l.x <- beta0 + beta1*x
theta.x <- l.x + k*x*dnorm(x,m=0,sd=s)
m.x <- 1/(1+exp(-theta.x))
logit.x <- 1/(1+exp(-l.x))
y <- rbinom(n,1,m.x)
# grid
nt<-101; tg<-seq(-3.5,3.5,length=nt)
l.t <- beta0 + beta1*tg
theta.t <- l.t + k*tg*dnorm(tg,m=0,sd=s)
m.t <- 1/(1+exp(-theta.t))
logit.t <- 1/(1+exp(-l.t))
op<-par(mfrow=c(1,2))
plot(tg,m.t,xlim=c(-3.5,3.5),ylim=c(0,1),type="l")
lines(tg,logit.t,col=2); points(x,y,col=4,pch=3)
plot(tg,theta.t,xlim=c(-3.5,3.5),type="l")
lines(tg,l.t,col=2)
par(op)
```

Write a script in R that fits a local logistic regression around $t_0=-0.5$. 
Then do the same for each element of vector `tg`. 

The pseudo-code should be like that:

1. Decide the bandwidth value: $h=0.35$, for instance.
2. For `t0= -0.5` do: 
    a. Define `x.t <- x - t0`.
    b. Compute the weight of each $x_i$ in `x` as $K((x_i-t_0)/h)$, where $K$ is a density function symmetric around 0 (use, for instance, `dnorm`).
    c. Fit the weighted generalized linear model `y ~ x.t` with family `binomial` and the weights computed before.
    d. Take the intercept coefficient $b_0$ of the preceding fitted generalized linear model and define the estimation $\hat{m}(t_0)$ of $m(t_0)$ as $\hat{m}(t_0)=1/(1+\exp(-b_0))$.  
3. Repeat the previous steps for each $t$ in `tg`.
4. Add the estimated function $\hat{m}(t)$, for $t\in$`tg`, 
to the graphic $(t,m(t))$, $t\in$`tg`.

\skip 

Estimating at $t_0$, for a given value of $t_0$:
```{r, message=FALSE, warning=FALSE}
t0 <- -.5
h <- .35
x.t <- x - t0
wt0 <- dnorm(x.t/h) 
glm.t0 <- glm(y~x.t,family = binomial, weights = wt0)
hat.beta0.t0 <- glm.t0$coefficients[1]
hat.m.t0 <- 1/(1+exp(-hat.beta0.t0)) 
```

Plotting the results:
```{r}
op<-par(mfrow=c(1,2))
plot(tg,m.t,xlim=c(-3.5,3.5),ylim=c(0,1),type="l")
lines(tg,logit.t,col=2); points(x,y,col=4,pch=3)
abline(v=t0,col=8,lwd=2)
points(x,.1*wt0/dnorm(0),col=8,pch=19)
points(t0,hat.m.t0,col=6,pch=19)
plot(tg,theta.t,xlim=c(-3.5,3.5),type="l")
lines(tg,l.t,col=2)
abline(v=t0,col=8,lwd=2)
points(x,min(l.t)+.1*(max(l.t)-min(l.t))*wt0/dnorm(0),col=8,pch=19)
points(t0,hat.beta0.t0,col=6,pch=19)
par(op)
```

Estimating at every `t` in `tg`:

```{r, message=FALSE, warning=FALSE}
h <- .35
hat.m.tg <- numeric(length(tg))
hat.beta0.tg <- numeric(length(tg))
for (j in (1:length(tg))){
  t <- tg[j]
  x.t <- x - t
  wt <- dnorm(x.t/h) 
  glm.t <- glm(y~x.t,family = binomial, weights = wt)
  hat.beta0.tg[j] <- glm.t$coefficients[1]
  hat.m.tg[j] <- 1/(1+exp(-hat.beta0.tg[j])) 
}
```

Plotting the results:

```{r}
op<-par(mfrow=c(1,2))
plot(tg,m.t,xlim=c(-3.5,3.5),ylim=c(0,1),type="l")
lines(tg,logit.t,col=2); points(x,y,col=4,pch=3)
abline(v=t0,col=8,lwd=2)
points(x,.1*wt0/dnorm(0),col=8,pch=19)
points(t0,hat.m.t0,col=6,pch=19)
lines(tg,hat.m.tg,col=6)
plot(tg,theta.t,xlim=c(-3.5,3.5),type="l")
lines(tg,l.t,col=2)
abline(v=t0,col=8,lwd=2)
points(x,min(l.t)+.1*(max(l.t)-min(l.t))*wt0/dnorm(0),col=8,pch=19)
points(t0,hat.beta0.t0,col=6,pch=19)
lines(tg,hat.beta0.tg,col=6)
par(op)
```

