---
title: 'Interpretability and Explainability in Machine Learning'
subtitle: 'Washington D.C. Bike Sharing Dataset'
author: "Pedro Delicado, Universitat Politècnica de Catalunya - Barcelona TECH"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
path.data.0 <- "./" #To be replaced by the appropriate path directory
```

## Include libraries
```{r, message=FALSE, warning=FALSE}
library(ranger)
library(randomForest)
library(caret)
library(vip)
library(DALEX)
library(DALEXtra)
library(lime)
library(iml)
library(localModel)
# library(fastshap) # Attention! It re-define "explain" from DALEX
if (require(ghostvar)){library(ghostvar)}
library(mgcv)
library(gridExtra)
```

# Washington D.C. Bike Sharing Dataset

[UC Irvine Machine Learning Repository, Bike Sharing Dataset](https://archive.ics.uci.edu/ml/datasets/bike+sharing+dataset). 


Fanaee-T and Gama (2014) compile two data sets with daily and hourly (respectively) information on the bike-sharing rental service in Washington D.C., corresponding to years 2011 and 2012.

> Fanaee-T, H. and J. Gama (2014). Event labeling combining ensemble detectors and background knowledge. *Progress in Artificial Intelligence*, **2**, 113-127.

The daily data set has 731 rows (one for each day of years 2011 and 2012, that was a leap year) and 16 variables.

   - `instant`:  row index, going from 1 to 731. 
   - `dteday`: date (character).
   - `mnth`:  1 for January, until 12 for December.
   - `season`: season (1:winter, 2:spring, 3:summer, 4:fall).
   - `yr`:  year (0: 2011, 1:2012).
   - `holiday`: weather day is holiday or not.
   - `weekday` : day of the week.
   - `workingday`: if day is neither weekend nor holiday is 1, otherwise is 0.  
   - `weathersit`: weather situation indicator:
		- 1: Clear, Few clouds, Partly cloudy, Partly cloudy.
		- 2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist.
		- 3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds.
		- 4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog.
   - `temp`:  temperature in Celsius. The values are divided to 41 (max).
   - `atemp`: feeling temperature in Celsius. The values are divided to 50 (max). 
   - `hum`: humidity. The values are divided to 100 (max).
   - `windspeed`: wind speed. The values are divided to 67 (max). 	
   - `casual`: count of rental bikes by casual users (not registered).
   - `registered`: count of rental bikes by registered users.
   - `cnt`: count of total rental bikes (casual + registered).

Potential response variables:

  - log of the number of total rental bikes.
  - log of the number of rental bikes by casual users.
  
	
```{r}
path.data <- paste0(path.data.0,"Bike-Sharing-Dataset/")
day <- read.csv(paste0(path.data,"day.csv"),as.is=TRUE)
hour <- read.csv(paste0(path.data,"hour.csv"),as.is=TRUE)
```


## Daily data (2011-2012)

### Show data

```{r}
head(day)
```

```{r}
#, fig.width=9, fig.height=6}
plot(day$cnt,col=day$season,pch=1+18*day$workingday,
     xlab="Day", ylab="Total users",
     main="Total daily users in 2011 and 2012, colored by season",
     ylim=c(0,9000))
legend("topleft",c("No working day","Working day"),pch=c(1,19),col=1)

```



```{r}
bikes <- day
# season: (1:winter, 2:spring, 3:summer, 4:fall)
bikes$spring <- as.integer(bikes$season==1)
bikes$summer <- as.integer(bikes$season==3)
bikes$fall <- as.integer(bikes$season==4)
#
bikes$mnth.2 <- bikes$mnth^2/12^2 
bikes$mnth.3 <- bikes$mnth^3/12^3 
bikes$mnth <- bikes$mnth/12 
# bikes$mnth <- (bikes$mnth-6.5)/5.5 
# bikes$mnth.2 <- bikes$mnth^2 
# bikes$mnth.3 <- bikes$mnth^3 
#
bikes$weather.good <- as.integer(bikes$weathersit==1)
bikes$weather.med <- as.integer(bikes$weathersit==2)
bikes$weather.bad <- as.integer(bikes$weathersit==3)

# [1] "instant"      "dteday"       "season"       "yr"          
# [5] "mnth"         "holiday"      "weekday"      "workingday"  
# [9] "weathersit"   "temp"         "atemp"        "hum"         
# [13] "windspeed"    "casual"       "registered"   "cnt"         
# [17] "spring"       "summer"       "fall"         "mnth.2"      
# [21] "mnth.3"       "weather.good" "weather.med"  "weather.bad" 

bikes2 <- bikes[,c(1, 14:16,4,17:19,5,20,21,8,23:24,10:13)]
# [1] "instant"     "casual"      "registered"  "cnt"         "yr"         
# [6] "spring"      "summer"      "fall"        "mnth"        "mnth.2"     
# [11] "mnth.3"      "workingday"  "weather.med" "weather.bad" "temp"       
# [16] "atemp"       "hum"         "windspeed"  
bikes2$log.casual <- log(bikes2$casual)
bikes2$log.cnt <- log(bikes2$cnt)
```



```{r}
plot(bikes2$cnt,col=bikes$season,pch=1+18*bikes$workingday,
     xlab="Day", ylab="Total users",
     main="Total users in 2011 and 2012, colored by season",
)
legend("topleft",c("No working day","Working day"),pch=c(1,19),col=4)
# abline(v=600.5,col=8,lwd=3)
# text(600,3550,"Training sample",pos=2)
# text(601,3550,"Test sample",pos=4)
```


```{r}
plot(bikes2$casual,col=bikes$season,pch=1+18*bikes$workingday,
     xlab="Day", ylab="Casual users",
     main="Casual users in 2011 and 2012, colored by season",
     ylim=c(0,3600))
legend("topleft",c("No working day","Working day"),pch=c(1,19),col=4)
# abline(v=600.5,col=8,lwd=3)
# text(600,3550,"Training sample",pos=2)
# text(601,3550,"Test sample",pos=4)
```


```{r}
plot(bikes2$log.cnt,col=bikes$season,pch=1+18*bikes$workingday,
     xlab="Day", ylab="log(cnt)",
     main="Total count of users (in logs) in 2011 and 2012, colored by season")
legend("bottomleft",c("No working day","Working day"),pch=c(1,19),col=4)
# abline(v=600.5,col=8,lwd=3)
# text(600,3,"Training sample",pos=2)
# text(601,3,"Test sample",pos=4)
```
```{r}
plot(bikes2$log.casual,col=bikes$season,pch=1+18*bikes$workingday,
     xlab="Day", ylab="log(casual users)",
     main="Casual users (in logs) in 2011 and 2012, colored by season")
legend("bottomleft",c("No working day","Working day"),pch=c(1,19),col=4)
# abline(v=600.5,col=8,lwd=3)
# text(600,3,"Training sample",pos=2)
# text(601,3,"Test sample",pos=4)
```

```{r}
plot(bikes2$temp, bikes2$cnt,col=day$yr+1,pch=19)
legend("topleft",c("Year 2011", "Year 2012"),col=c(1,2),pch=19)
```


### Data processing: Creating training and test sets, and removing certain variables
```{r}
# names(bikes2)
#  [1] "instant"     "casual"      "registered"  "cnt"        
#  [5] "yr"          "spring"      "summer"      "fall"    
#  [9] "mnth"        "mnth.2"      "mnth.3"      "workingday" 
# [13] "weather.med" "weather.bad" "temp"        "atemp"      
# [17] "hum"         "windspeed"   "log.casual"  "log.cnt"

day_selected <- bikes2[,c(4:15,17,18)] # select columns 4-14 and 17 and 18
head(day_selected)

I.train <- sort( sample(1:dim(day_selected)[1],500,replace=FALSE) ) # vector of 500 randomly sampled indices from the row indices of day_selected sorted in ascending order
day_sel_train <- day_selected[I.train,]
day_sel_test <- day_selected[-I.train,]
```

# 1. Fit a Random Forest

## (1) Split-criterion in a regression tree.
Let us revise the computation of the split-criterion in a regression tree.

```{r}
split_criterion <- function(y, split.var=y){
  y <- y[order(split.var)]
  n <- length(y)
  uni.spl.var <- sort(unique(split.var))
  m <- length(uni.spl.var)
  Q <- numeric(m+1)
  Q[1] <- Q[m+1] <- (n-1)*var(y)
  #Q[2] <- (n-2)*var(y[2:n])
  for (i in 2:m){
    ni <- sum(split.var<uni.spl.var[i])
    Q[i] <- 
      ifelse(ni>1,(ni-2)*var(y[1:ni]),0) + 
      ifelse(n-ni>1, (n-ni)*var(y[(ni+1):n]),0)
  }
  #Q[m] <- (n-2)*var(y[1:(n-1)])
  Q <- Q/Q[1]
  wminQ <- which.min(Q)
  return(list(Q=Q, split.var=split.var, 
              split=mean(uni.spl.var[wminQ+c(-1,0)]),
              Qsplit=Q[wminQ],
              wminQ=wminQ/(m+1))
         )
}
```

```{r}
aux <- split_criterion(day_sel_train$cnt)
Q0<-aux$Q
m <- length(Q0)-1
plot((0:m)/m,Q0, type="l", lty=2, 
     xlab="Cumul. freq. of split variables",ylab="Split criterion")
abline(h=1,col=8,lty=2)

p <- dim(day_sel_train)[2]-1
splits <- numeric(p)
Qsplits <- numeric(p)+1
wminQ <- numeric(p)
for (j in (1:p)[-(6:7)]){
  aux <- split_criterion(day_sel_train$cnt,split.var = day_sel_train[,j+1])
  Q<-aux$Q
  #print(Q)
  m <- length(Q)-1
  splits[j] <- aux$split
  Qsplits[j] <- aux$Qsplit
  wminQ[j]<-aux$wminQ
  lines((0:m)/m,Q, type="l", col=j, lty=j)
}
var.split <- which.min(Qsplits)
abline(v=wminQ[var.split],col=var.split, lty=var.split)
text(wminQ[var.split],(min(Q0)+Qsplits[var.split])/2,
     names(day_sel_train)[var.split+1],col=var.split,pos=4)

# every curve corresponds to one explanatory variable
# x axis is every possible value of splits
# best cut s for month would be o.25 (cyan)
# 
```


## (2) Random forest. Variable Importance by the reduction of the impurity.
First the *Variable Importance* is measured by the reduction of the **impurity** at the splits defined by each variable.
The *impurity* measure is the variance of the responses for regression (and the Gini index for classification).

```{r}
model_rf_imp <- ranger(
  cnt ~ .,
  data = day_sel_train[,-c(7,8)], 
  importance='impurity' # default = none, no importances will be computed
)
print(model_rf_imp)
# ranger.importance -> survival = time to a specific event
```


## Random forest (3). Variable Importance by out-of-bag random permutations.
Now the measure of *Variable Importance* by *random permutations* of the values of each variable in the out-of-bag samples.

```{r}
model_rf_perm <- ranger(
  cnt ~ .,
  data = day_sel_train[,-c(7,8)], 
  importance='permutation'
)
print(model_rf_perm)
```

## Graphical representation of Variable Importance
```{r}
# vip = variable importance
rf_imp_vip <- vip(model_rf_imp, num_features = 11)
rf_perm_vip <- vip(model_rf_perm, num_features = 11)
grid.arrange(rf_imp_vip, rf_perm_vip, ncol=2, top="Left: Reduction in impurity at splits. Right: Out-of-bag permutations")
# plots 11 most important variables (default is 10)
```


# 2. Shapley Values

```{r}
rf_imp_vip <- vip(model_rf_imp, num_features = 11)
rf_perm_vip <- vip(model_rf_perm, num_features = 11)
rf_shapley <- vip(model_rf_imp, method="shap",
                  pred_wrapper=yhat, num_features = 11,
                  train = day_sel_train, # argument 'train' must be provided
                  newdata=day_sel_test[,-c(1,7,8)]) # exclude columns 1, 7, 8 
grid.arrange(rf_imp_vip, rf_perm_vip, rf_shapley,
             ncol=2, nrow=2,
             top="Top left: Impurity. Top right: oob permutations. Bottom left: Shapley values"
            )
```



## Fitting a linear model and a gam model

```{r}
lm_bikes <- lm(cnt ~ ., data = day_sel_train)
(summ_lm_bikes <- summary(lm_bikes))
```

```{r}
gam_bikes <- gam(cnt ~ yr + spring + summer + fall + s(mnth) + 
                   workingday + weather.med + weather.bad + 
                 s(temp) + s(hum) + s(windspeed), 
                 data = day_sel_train)
(summ_gam_bikes <- summary(gam_bikes))
```

```{r}
plot(gam_bikes)
```

```{r}
#DALEX::explain(lm_bikes, X=day_sel_train,
#        pred_wrapper=predict.lm,
#        newdata=day_sel_test[,-11])

lm_bikes_shapley <- vip(lm_bikes, method="shap",
                  pred_wrapper=predict.lm,
                  train=day_sel_train, # train set must be specified
                  newdata=day_sel_test[,-1], 
                  num_features = 14,
                  exact=TRUE)

plot(lm_bikes_shapley)
```

```{r}
gam_bikes_shapley <- vip(gam_bikes, method="shap",
                  pred_wrapper=predict.gam,
                  train=day_sel_train, # train set must be specified
                  newdata=day_sel_test[,-1],
                  num_features = 11,
                  exact=TRUE)

plot(gam_bikes_shapley)
```


```{r}
grid.arrange(rf_imp_vip, rf_shapley, 
             lm_bikes_shapley, gam_bikes_shapley, 
             ncol=2, nrow=2,
             top="1,1: RF Impurity. 1,2: Shapley RF. 2,1: Shapley lm. 2,2: Shapley gam"
)
```

# 3. Relevance by Ghost Variables

```{r,fig.width=8,fig.height=12}
library(grid)
source("relev.ghost.var.R")
Rel_Gh_Var <- relev.ghost.var(model=gam_bikes, 
                              newdata = day_sel_test[, -1],
                              y.ts = day_sel_test[, 1],
                              func.model.ghost.var = lm
)
plot.relev.ghost.var(Rel_Gh_Var,n1=500,ncols.plot = 3)
# year and temp are the most relevant variables
```

```{r,fig.width=8,fig.height=6}
aux <- cbind(Rel_Gh_Var$relev.ghost,gam_bikes_shapley$data$Importance)
plot(aux[,1],aux[,2],col=0,xlab="Relev. by Ghost Variables",ylab="Shapley Var. Imp.")
text(aux[,1],aux[,2],row.names(aux))
```

# 4. Global Importance Measures and Plots using the library DALEX

```{r}
# Using library DALEX
# help(package="DALEX")
# ? explain
explainer_rf <- explain.default(model = model_rf_imp,  
                               data = day_sel_test[, -c(1,7,8)],
                               y = day_sel_test$cnt, 
                               label = "Random Forest")

# first thing we do in dalex is creating a interface for all different explanation methods to fitted model  = explainer
# with this explainer created you can use everything in Dalex
# it is the intermediate step needed to use the fitted model in dalex
# the explainer must be able to extract the function used for predictions
``` 

## Random Permutations Variable Importance

```{r}
# ?model_parts
Rnd_Perm <- model_parts(
  explainer_rf,
  N = NULL, # All available data are used
  B = 10   # number of permutations to be used, with B = 10 used by default
)

Rnd_Perm

plot(Rnd_Perm)
```

```{r}
# Creating an object of class "vi" from the "model_parts" object created by DALEX # This is necesarry to 
aux.plot <- plot(Rnd_Perm)
dropout_loss.y <- Rnd_Perm$dropout_loss[1]
aux.I <- order(-aux.plot$data$dropout_loss.x)
rf_perm_DALEX_as_vi <- tibble::tibble(aux.plot$data[aux.I,c(2,4)])
class(rf_perm_DALEX_as_vi) <- c("vi", class(rf_perm_DALEX_as_vi))
names(rf_perm_DALEX_as_vi) <- c("Variable", "Importance")
rf_perm_DALEX_as_vi$Importance <- 
  rf_perm_DALEX_as_vi$Importance - dropout_loss.y

# Creating the ggpolt: 
rf_perm_DALEX_vip <- vip(rf_perm_DALEX_as_vi)

grid.arrange(rf_imp_vip, rf_perm_vip,
             rf_perm_DALEX_vip, ncol=2, nrow=2,
             top="Top left: Impurity. Top right: oob permutations. Bottom left: test sample permutations"
             )
```


## Partial Dependence Plot

The function that allows computation of PD profiles in the DALEX package is 
`model_profile()`. 
The only required argument is `explainer`, which indicates the explainer-object 
(created with the `DALEX::explain()` function) for the model to be explained.

```{r, fig.height=12, fig.width=6}
# ?model_profile
PDP_rf <- model_profile(
  explainer=explainer_rf,
  variables = NULL,  # All variables are used
  N = NULL, # All available data are used
  groups = NULL,
  k = NULL,
  center = TRUE,
  type = "partial" #  partial, conditional or accumulated
)

plot(PDP_rf, facet_ncol=2)
```

## Local (or Conditional) Dependence Plot

```{r, fig.height=12, fig.width=6}
CDP_rf <- model_profile(
  explainer=explainer_rf,
  variables = NULL,  # All variables are used
  N = NULL, # All available data are used
  groups = NULL,
  k = NULL,
  center = TRUE,
  type = "conditional" #  partial, conditional or accumulated
)

plot(CDP_rf, facet_ncol=2)
```


## Accumulated Local Effects

```{r, fig.height=12, fig.width=6}
# ?model_profile
ALE_rf <- model_profile(
  explainer=explainer_rf,
  variables = NULL,  # All variables are used
  N = NULL, # All available data are used
  groups = NULL,
  k = NULL,
  center = TRUE,
  type = "accumulated" #  partial, conditional or accumulated
)

plot(ALE_rf, facet_ncol=2)
```


# 5. Local explainers with library DALEX

We choose an instance in of 2011, and another one in 2012, the prediction for which we want to explain.

```{r}
aux <- which(day[-I.train,2]=="2011-10-04")
(october_4_2011 <- day_sel_test[aux,])

aux <- which(day[-I.train,2]=="2012-10-04")
(october_4_2012 <- day_sel_train[aux,])
```

## SHAP 

```{r}
bd_rf <- predict_parts(explainer = explainer_rf,
                 new_observation = october_4_2011,
                            type = "shap") # based on shapely value

bd_rf
plot(bd_rf)
```

```{r}
bd_rf <- predict_parts(explainer = explainer_rf,
                 new_observation = october_4_2012, # next year
                            type = "shap")

bd_rf
plot(bd_rf)
# all variables have a positive contribution to prediction compared to average prediction
```

## Break-down plots

```{r}
bd_rf <- predict_parts(explainer = explainer_rf,
                 new_observation = october_4_2011,
                            type = "break_down") # using break down plots

bd_rf
plot(bd_rf, max_features=11)

# intercept = the average of all predictions = 4459.909
# prediction (bar) = difference: sum of difference of all variable is the distance
# = distribute distance among all predictor variables
```

```{r}
bd_rf <- predict_parts(explainer = explainer_rf,
                 new_observation = october_4_2012,
                            type = "break_down")

bd_rf
plot(bd_rf, max_features=11)

# distance is positive and very large
# prediction = distance between average prediction and actual prediction
# we distribute the distance between all predictor vars
```


## LIME

```{r} 
lime_rf <- predict_surrogate(explainer = explainer_rf, 
                  new_observation = october_4_2011[,-1], 
                  type = "localModel")
                  #type = "iml") # it does not work
                  #n_features = 6, 
                  #n_permutations = 1000,
                  #type = "lime") # it does not work

lime_rf 

plot(lime_rf)
```

```{r} 
lime_rf <- predict_surrogate(explainer = explainer_rf, 
                  new_observation = october_4_2012[,-1], 
                  type = "localModel")
                  #type = "iml") # it does not work
                  #n_features = 6, 
                  #n_permutations = 1000,
                  #type = "lime") # it does not work

lime_rf 

plot(lime_rf)
```


## Local Graphics

### Individual conditional expectation (ICE) plot, or ceteris paribus plot

```{r, fig.height=12, fig.width=6}
cp_rf <- predict_profile(explainer = explainer_rf, 
                           new_observation = october_4_2011)
cp_rf

plot(cp_rf, facet_ncol=2)
```

```{r, fig.height=12, fig.width=6}
cp_rf <- predict_profile(explainer = explainer_rf, 
                           new_observation = october_4_2012)
cp_rf

plot(cp_rf, facet_ncol=2)
```

```{r}
mp_rf <- model_profile(explainer = explainer_rf,
  variables = "temp",
  N = 100,
  type = "partial"
)

plot(mp_rf, geom = "profiles") +
  ggtitle("Ceteris-paribus and partial-dependence profiles for temp")

# grey = all individual expectation plots
# blue = average
```