---
title: "Example of ISOAMP algorithm"
author: "Pedro Delicado"
date: "`r format(Sys.time(), '%d de %B de %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Portion of S shape curve, with radious R, of length 2*pi with noise

t <- seq(-1.5*pi,1.5*pi,l=100)
R<- 1
n<-75
sd.residuals <- .15
```

```{r fig.height=5, fig.width=7}
#windows(width=7,height=5)
set.seed(1)
A <- 2.6
plot(0,0,type="n",xlab="",ylab="",axes=FALSE,
     #     main="Ilustracion Paso 1 de ISOMAP",
     xlim=c(-1-2*sd.residuals,1+2*sd.residuals + 2*A),
     ylim=c(-2-2*sd.residuals,2+2*sd.residuals))

# S plot with points
y <- R*sign(t) - R*sign(t)*cos(t/R)
x <- -R*sin(t/R)
lines(x,y,col=2,lwd=2)
rt <- sort(runif(n)*3*pi - 1.5*pi)
residuals <- rnorm(n)*sd.residuals
ry <- R*sign(rt) - (R+residuals)*sign(rt)*cos(rt/R)
rx <- -(R+residuals)*sin(rt/R)
points(rx,ry,col=1,pch=19,cex=.6)
#lines(x,y,col=2,lwd=2)

# S plot with points and all the connections
temp.seed <- .Random.seed
for (i in 1:(n-1)){
  for (j in (i+1):n){
    if ( runif(1) <= .1 ) lines(A + rx[c(i,j)],ry[c(i,j)],col=3)
  }
}
lines(A+x,y,col=2,lwd=2)
points(A + rx,ry,col=1,pch=19,cex=.6)

# S plot with points, with connections <= eps
eps <- .5
XY <- cbind(rx,ry)
lines(2*A+x,y,col=2,lwd=2)
for (i in 1:(n-1)){
   for (j in (i+1):n){
       if ( dist(XY[c(i,j),])<= eps )lines(2*A + rx[c(i,j)],ry[c(i,j)],col=3)
   }
}
points(2*A + rx,ry,col=1,pch=19,cex=.6)
text(2*A,2.35,"epsilon=0.5")
#dev.off()
```

```{r}
library(vegan)
ismp <- isomap(dist(XY),eps=eps)
```

```{r fig.height=3, fig.width=9}
# windows(width=9,height=3)
# pdf("ej_isomap_S_res_Engl.pdf",width=9,height=3,paper="special")
aux.plot <- plot(ismp,n.col=3,main="Output of ISOMAP Algorithm")
points(aux.plot,"sites",pch=19,cex=.6)
#dev.off()
```


It seems that $q=1$ can be enough:
```{r}
q<-1 # 1-dim config
ismp.1 <- isomap(dist(XY),eps=eps, ndim=q)

lambda <- ismp.1$points[,1] # 1-dimemsional configuration
plot(lambda,rep(0,n),#as=1,    
     xlab="lambda (1-dimemsional configuration)",
     ylab="")
if (ismp.1$criterion=="epsilon"){
  title(main=paste0("ISOMAP, epsilon=",eps))
}else{
  title(main=paste0("ISOMAP, k=",k))
}
text(lambda,rep(0,n),1:n,pos=3,cex=.4)
```

The 1-dimensional configuration $\lambda$ given by ISOMAP is a good recovery of the true underline 1-dimensional generating variable `rt`. The linear relation between $\lambda$ and `rt` is almost perfect:

```{r,fig.asp=1}
plot(lambda,rt, as=1)
```

Now we plot the 1-dimensional structure of the data obtained from ISOMAP using smoothing techniques:

```{r,fig.height=8}
#install.packages("mgcv") # To perform spline smoothing
library(mgcv)
smooth.x <- gam(rx~s(lambda))
smooth.y <- gam(ry~s(lambda))

s.x <- smooth.x$fitted.values
s.y <- smooth.y$fitted.values

op <- par(mfrow=c(2,1))
plot(lambda,rx,col=8, pch=19, cex=1, main="x versus lambda")
lines(lambda, s.x, col=1, lwd=2)
plot(lambda,ry,col=8, pch=19, cex=1, main="y versus lambda")
lines(lambda, s.y, col=1, lwd=2)
par(op)
```
We add the nonlinear curve fitted by ISOMAP to the 2-dimensional representation of the original data:

```{r}
A <- 2.6
plot(0,0,type="n",xlab="",ylab="",axes=FALSE,
     xlim=c(-1-2*sd.residuals,1+2*sd.residuals + 2*A),
     ylim=c(-2-2*sd.residuals,2+2*sd.residuals))

# S plot with points
lines(x,y,col=2,lwd=2)
points(rx,ry,col=8,pch=19,cex=.8)
lines(s.x,s.y,col=1,lwd=2)

legend(.75*A,.75*A,c("Observed points","True generating curve", "Estimated curve"),
       col=c(8,2,1), pch=c(19,NA,NA), lwd=c(NA,2,2), bty="n", cex=.8)
```
