---
title: "Density based clustering: DBSCAN"
author: "Pedro Delicado"
date: "`r format(Sys.time(), '%d/%b/%Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r global_values}
save_pdf<-FALSE # TRUE
width_pdf <- 6
height_pdf <- 6
```

## 1. DBSCAN concepts

(number of clusters must not be specified in advance)

### A simulated data example
We simulate 2-dimensional data clustered in 2 groups with additional outliers.

```{r}
set.seed(1234)

n1 <- 50
n2 <- 20
nout <- 10

# Cluster 1: 
theta <- seq(-pi/4,3*pi/4,length=n1)
s.eps <- .1
eps <- matrix(s.eps*rnorm(2*n1),ncol=2)
X1 <- cbind(cos(theta),sin(theta))+eps

# Cluster 2:
s2 <- .2
X2 <- matrix(s2*rnorm(2*n2),ncol=2)

# Outliers:
theta.out <- runif(nout,0,2*pi)
s.eps.out <- .2
eps.out <- matrix(s.eps*rnorm(2*nout),ncol=2)
R.out <- 1.5
Xout <- R.out*cbind(cos(theta.out),sin(theta.out))+eps.out

# Final data set:
X <- rbind(X1,X2,Xout)
cl.true <- c(rep(1,n1), rep(2,n2), rep(0,nout))
n<-n1+n2+nout
```

```{r,fig.asp=1,fig.width=6}
if (save_pdf) pdf("data.pdf",width = width_pdf, height = height_pdf)
plot(X, asp=1,xlab="x",ylab="y",xlim=c(-2,2),ylim=c(-2,2))
if (save_pdf) dev.off()
```

```{r,fig.asp=1,fig.width=6}
if (save_pdf) pdf("data_cl.pdf",width = width_pdf, height = height_pdf)
plot(X, col=1+cl.true,asp=1,xlab="x",ylab="y",xlim=c(-2,2),ylim=c(-2,2))
if (save_pdf) dev.off()
```


### $\varepsilon$-neigborhoods:

```{r}
epsilon <- .25
minPts <- floor(n/20)
```


```{r,fig.asp=1,fig.width=6}
if (save_pdf) pdf("eps_neig.pdf",width = width_pdf, height = height_pdf)
symbols(x=X[,1],y=X[,2],circles=rep(epsilon,n), inches=FALSE, fg=1, bg=8, 
        asp=1,xlim=c(-2,2),ylim=c(-2,2),
        main=paste("epsilon=",epsilon),xlab="x",ylab="y")
symbols(x=X[,1],y=X[,2],circles=rep(epsilon,n), inches=FALSE, fg=1, bg=NA,
        add=TRUE)
points(X,pch=19)
if (save_pdf) dev.off()
```

### Core points: 

```{r,fig.asp=1,fig.width=6} 
if (save_pdf) pdf("core.pdf",width = width_pdf, height = height_pdf)
D <- as.matrix(dist(X))
core.pts <- which(apply(D<=epsilon, 1, sum)>=minPts)
no.core.pts <- (1:n)[-core.pts]
n.core <- length(core.pts)
n.no.core <- n-n.core

symbols(x=X[core.pts,1],y=X[core.pts,2],circles=rep(epsilon,n.core), inches=FALSE, fg=2, bg="pink",
        asp=1,xlim=c(-2,2),ylim=c(-2,2),
        main=paste("epsilon=",epsilon,", minPts=",minPts),xlab="x",ylab="y")
symbols(x=X[core.pts,1],y=X[core.pts,2],circles=rep(epsilon,n.core), inches=FALSE, fg=2, bg=NA,
        add=TRUE)
symbols(x=X[no.core.pts,1],y=X[no.core.pts,2],circles=rep(epsilon,n.no.core), inches=FALSE, 
        fg=1, bg=NA,
        asp=1,xlim=c(-2,2),ylim=c(-2,2),add=TRUE)
points(X[core.pts,],pch=19,col=2)
points(X[no.core.pts,],pch=19,col=1)
legend("topright",c("Core points","No core points"),pch=19,col=c(2,1))
if (save_pdf) dev.off()
```

### Border points and outliers: 

```{r,fig.asp=1,fig.width=6}
if (save_pdf) pdf("border_outliers.pdf",width = width_pdf, height = height_pdf)
D.nc.c <- D[no.core.pts,core.pts]
border.pts <- no.core.pts[which(apply(D.nc.c,1,min)<=epsilon)]
outliers <- setdiff(no.core.pts,border.pts)
n.border<- length(border.pts)
n.outliers <- n.no.core - n.border

symbols(x=X[core.pts,1],y=X[core.pts,2],circles=rep(epsilon,n.core), inches=FALSE, fg=2, bg="pink",
        asp=1,xlim=c(-2,2),ylim=c(-2,2),
        main=paste("epsilon=",epsilon,", minPts=",minPts),xlab="x",ylab="y")
symbols(x=X[core.pts,1],y=X[core.pts,2],circles=rep(epsilon,n.core), inches=FALSE, fg=2, bg=NA,
        add=TRUE)
symbols(x=X[border.pts,1],y=X[border.pts,2],circles=rep(epsilon,n.border), inches=FALSE, 
        fg=4, bg=NA,add=TRUE)
symbols(x=X[outliers,1],y=X[outliers,2],circles=rep(epsilon,n.outliers), inches=FALSE, 
        fg=1, bg=NA,ylim=c(-2,2),add=TRUE)
points(X[core.pts,],pch=19,col=2)
points(X[border.pts,],pch=19,col=4)
points(X[outliers,],pch=19,col=1)
legend("topright",c("Core points","Border points","Outliers"),pch=19,col=c(2,4,1))
if (save_pdf) dev.off()
```

## 2. DBSCAN algorthim

```{r}
remain.pts <- setdiff(1:n,outliers)
clusters <- list()
cl.found <- numeric(n)
j <- 0
while (length(remain.pts)>0) {
  j <- j+1
  cluster.j <- sample(intersect(remain.pts,core.pts),1)
  remain.pts <- setdiff(remain.pts,cluster.j)
  D.aux <- D[remain.pts,cluster.j]
  eps.reachable <- remain.pts[D.aux<=epsilon]
  while (length(eps.reachable)>0){
    cluster.j <- union(cluster.j,eps.reachable)
    remain.pts <- setdiff(remain.pts,cluster.j)
    if (length(remain.pts)>0){
      D.aux <- matrix(D[remain.pts,cluster.j],ncol=length(cluster.j))
      eps.reachable <- remain.pts[which(apply(D.aux,1,min)<=epsilon)]
    }else{
      eps.reachable <- numeric(0)
    }
  }
  clusters[[j]]<-sort(cluster.j)
  cl.found[clusters[[j]]] <- j
}

table(cl.true, cl.found)
``` 

```{r,fig.asp=1,fig.width=6}
if (save_pdf) pdf("my_dbscan.pdf",width = width_pdf, height = height_pdf)
fg.col <- c("red","darkgreen","blue")
bg.col <- c("pink","green","lightblue")
plot(X[outliers,],pch=19,col=1,
     asp=1,xlim=c(-2,2),ylim=c(-2,2),
     main=paste("epsilon=",epsilon,",minPts=",minPts),
     xlab="x",ylab="y")
J <- length(clusters)
for (j in (1:J)){
  symbols(x=X[clusters[[j]],1],y=X[clusters[[j]],2],
          circles=rep(epsilon,length(clusters[[j]])), 
          inches=FALSE, fg=fg.col[j], bg=bg.col[j], add=TRUE)
  points(X[clusters[[j]],],pch=19,col=fg.col[j])
}
points(X[outliers,],pch=19,col=1)
legend("topright",c(paste("Cluster ",1:J),"Outliers"),
       pch=19,col=c(fg.col[1:J],1))
if (save_pdf) dev.off()
```

## 3. DBSCAN implementations in R

### Library `fpc`

`fpc::dbscan` generates a density based clustering of arbitrary shape as introduced in Ester et al. (1996).

```{r,fig.asp=1,fig.width=6}
if (save_pdf) pdf("fpc_dbscan.pdf",width = width_pdf, height = height_pdf)
library(fpc)
# ?fpc::dbscan
fpc.ds <- fpc::dbscan(X,eps = epsilon, MinPts = minPts, showplot = 0)
plot(fpc.ds,X, main=paste("fpc::dbscan; epsilon=",epsilon,",minPts=",minPts),
     xlab="x",ylab="y",asp=1,xlim=c(-2,2),ylim=c(-2,2))
if (save_pdf) dev.off()
```

```{r}
print(fpc.ds)
table(cl.true, fpc.ds$cluster)
```

### Library `dbscan`

`dbscan::dbscan`: Fast reimplementation of the DBSCAN clustering algorithm using a $k$-dimensional tree (kd-tree).

```{r,fig.asp=1,fig.width=6}
if (save_pdf) pdf("dbscan_dbscan.pdf",width = width_pdf, height = height_pdf)
library(dbscan)
# ?dbscan::dbscan
dbscan.ds <- dbscan::dbscan(X,eps = epsilon, minPts = minPts)
plot(X, col=dbscan.ds$cluster+1, pch=dbscan.ds$cluster+1, 
     main=paste("dbscan::dbscan; epsilon=",epsilon,",minPts=",minPts),
     xlab="x",ylab="y",asp=1,xlim=c(-2,2),ylim=c(-2,2))
if (save_pdf) dev.off()
```

```{r}
table(cl.true, dbscan.ds$cluster)
```